Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-04-22 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
update

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S_{ua})}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. Boundary conditions also include no-flux across upper and lower borders. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial permeability distribution creates a rough pressure field, \ie pressure gradient is discontinuous across the interface between different permeability zones, thus the fluid flows to the larger permeability region. %Figure \ref{fem_cv_represent_a}(b-c)describe the saturation field at different times.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Studies by \citet{langtangen_1992} on fluid flow dynamics in heterogeneous porous media (analytical and numerical solutions) demonstrated that hyperbolic Buckley-Leverett model is inherently unstable, \ie the expected uniform interface front collapses as any perturbation in physical parameters are imposed to the problem.

Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

\medskip
Here, two numerical simulations were conducted with fixed and adaptive mesh resolutions to qualitatively assess model capability to capture fingers dynamics. In order to trigger the formation of fingers, regions with sharp permeability gradient were introduced. The computational domain, Fig.~\ref{fig:testcase_heter_domain}a, consists of a rectangular geometry fully saturated with a fluid (except by a squared region containing 50$\%$ of a second fluid -- wetting fluid phase, Fig.~\ref{fig:testcase_heter_domain}b) and divided into 5 regions with prescribed permeability distribution. A no-flux boundary condition was imposed across upper and lower borders, whilst pure wetting fluid phase $\left(\text{\ie} S_{1}=1\right)$ is driven into the domain from the left-hand side face at velocity $u_{1}=1$. Viscosity ratio was set to 10.

>>>>

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha}\hspace{1cm} \\ %\text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S_{u,{\alpha}}}$ is the source term %associated with the Darcy and continuity equations.
which may include gravity and/or capillarity, respectively. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.

@@

%%%%%%%%%%%% 
\subsection{Initial Model Validation}\label{section:results_initial_model_validation}
Numerical simulations \citep[based on lab experiments due to][]{evans_1994,dawe_2008} were conducted to demonstrate the model capability to capture viscous crossflow during immiscible displacement in heterogeneous porous media. The 2D domain, shown in Fig.~\ref{fem_cv_represent_a}, is $4$ $\times$ $2$ unit-length and fully saturated with fluid 2 (VR=1). Fluid 1 is injected from the left-hand side of the domain with constant velocity of $u=1$. \green{No normalflux boundary conditions are implied across the upper and lower borders}. The domain consists of four regions in which each quarter is represented by a permeability value, Fig.~\ref{fem_cv_represent_a}(a). The spatial permeability distribution creates a rough pressure field, \ie pressure gradient is discontinuous across the interface between different permeability zones, thus the fluid flows to the larger permeability region. %Figure \ref{fem_cv_represent_a}(b-c)describe the saturation field at different times.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Studies by \citet{langtangen_1992} on fluid flow dynamics in heterogeneous porous media (analytical and numerical solutions) demonstrated that hyperbolic Buckley-Leverett model is inherently unstable, \ie the expected uniform interface front collapses as any perturbation in physical parameters are imposed to the problem. Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

\medskip
Here, two numerical simulations were conducted with fixed and adaptive mesh resolutions to qualitatively assess model capability to capture fingers dynamics. In order to trigger the formation of fingers, regions with sharp permeability gradient were introduced. The computational domain, Fig.~\ref{fig:testcase_heter_domain}a, consists of a rectangular geometry fully saturated with a fluid (except by a squared region containing 50$\%$ of a second fluid -- wetting fluid phase, Fig.~\ref{fig:testcase_heter_domain}b) and divided into 5 regions with prescribed \green{synthetic permeability distribution with values between $1$ and $5$}. A no-flux boundary condition was imposed across upper and lower borders, whilst pure wetting fluid phase $\left(\text{\ie} S_{1}=1\right)$ is driven into the domain from the left-hand side face at velocity $u_{1}=1$. Viscosity ratio was set to 10.
